---
title: "reshaping_data"
output: html_document
date: "2023-03-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

first to load in the libraries you will use
```{r load libraries}
library(tidyverse)
library(dplyr)
```

now to load in the data we are using
```{r}
meta_data_fp <- "C:/Users/c76210ml/OneDrive - The University of Manchester/Desktop/coffee_n_coding/data/metatable-master.xlsx"

meta_data <- readxl::read_xlsx(meta_data_fp)

data_fp <- "C:/Users/c76210ml/OneDrive - The University of Manchester/Desktop/coffee_n_coding/data/metadata.tsv"

data <- readr::read_tsv(data_fp)

```

Now that the data is loaded in you can see how it looks by looking at the top right window of your screen

you'll see a variable called 'data', which is your sequencing data, and one called 'meta_data', which is the conditions for your experiments

if you click on 'data', it will open a tab that looks like an excel spreadsheet
additionally, you can click the arrow next to the variable name in the Environment screen (top right) and it will list the variable names for you and a preview of the data

Now we can start to reshape the data for your purposes

```{r reshape data, combine meta and data tables}

meta_data <- meta_data[-1,]

colnames(meta_data)[2] <- 'condition'

meta_data <- meta_data %>%  pivot_wider(names_from = 'SampleID', values_from = 'condition') 

meta_data <- janitor::clean_names(meta_data)

meta_data<- meta_data %>% 
  pivot_longer(everything())

data <- data[-1,] 

#renaming the columsn F2 Control... etc to replace the space with an underscore for manipulation purposes
data <- janitor::clean_names(data)

data <- data %>% 
  pivot_longer(cols = dv1:f5_control)%>% 
  left_join(meta_data, by= 'name')

colnames(data)[6] <- 'hits_to_bacteria_sequence'
colnames(data)[7] <- 'condition'
```

now if you look at our 'data' you can see that there is a new column called condition that gives whether a hit was in the 'soil','control' or 'negative' conditions

```{r adding in total_reads, sum_ratio, reads_ratio}

data <- data %>% 
  separate(taxon, sep = '; ', c('domain','phylum','class', 'order', 'family', 'genus', 'species'))
  
data <- data %>% 
  group_by(id) %>% 
  mutate(total_reads = sum(as.numeric(hits_to_bacteria_sequence))) 

sum_controls <- data %>% 
  group_by(id) %>%  
  filter(!(condition =='soil')) %>% 
  mutate(sum_controls = sum(as.numeric(hits_to_bacteria_sequence)))

sum_controls_to_add <- sum_controls %>% 
  select(id, condition, sum_controls)

data <- data %>% 
  left_join(sum_controls_to_add)

```

